---
title: "Statistical inference with the GSS data"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
  pdf_document: default
---

## Setup
### Load packages

```{r load-packages, message = FALSE}
#devtools::install_github("haleyjeppson/ggmosaic")
#install.packages('vcd')
#install.packages('pastecs')
#install.packages(c("ggplot2", "ggpubr", "tidyverse", "broom", "AICcmodavg"))
#install.packages('tinytex')
#tinytex::install_tinytex()

library(ggplot2)
library(dplyr)
library(statsr)
library(readr)
library(ggmosaic)
library(vcd)
library(pastecs)
library(skimr)
library(ggpubr)
library(tidyverse)
library(broom)
library(AICcmodavg)
library(car)
library(plyr)
```

### Load data

```{r load-data}
load("gss.Rdata")
```

* * *

### Part 1: Data

### How are the observations in the sample collected?
According to the official GSS website, the GSS is a "nationally representative" survey of adults in the United States. 
It utilizes an "area probability design that randomly selects respondents in households across the United States."

Due to this, it's safe to assume that the data collected is representative of the population of adults at large. However, it's likely that due to the nature of data collection, causal connections cannot be inferred, but the results can be generalized.

* * *

### Part 2: Research questions

### Investigating the data
#### Potentially interesting variables to investigate further
1. Religiosity (relig) and Education (degree). Relationship between level of education and religious affiliation?
2. Religiosity (relig) and Political Party (partyid). Relationship between religiosity and political party? ***
3. Religiosity (relig), Political Party (partyid) and Education level (degree).
4. Is level of education (degree) and family income (coninc) associated? ***
5. News and overall happiness.
6. What is the average # of Children adults have.
7. What is the average age of adults in the US?

```{r}
#### Investigate the dataset

#head(gss)
#colnames(gss)
#names(gss)
#str(gss)
#glimpse(gss)
#skim_without_charts(gss)
#summary(gss)
```

* * *

## Analysis #1 - Religion and Party Affiliation (Chi-Square Test of Independence)

### Part 3: Exploratory data analysis

Does there appear to be a relationship between religiosity and political party 
affiliation?

The parties contained in this analysis are Democrat, Independent, Republican
and Other.

For the purposes of this analysis, anyone affiliated with a religion is considered
religious. Those that are nearly or strongly leaning toward a particular political
party will be considered associated with that party.

#### Initial Analysis and Cleaning
```{r}
##### Keep specific columns for analysis.
trimmed_gss_analysis_one <- gss %>%
  select(c(relig, partyid))

##### Check for N/A's.
sapply(trimmed_gss_analysis_one, function(x) sum(is.na(x)))

##### Remove rows with N/A' (for this analysis, it was acceptable.)
trimmed_gss_analysis_one <- na.omit(trimmed_gss_analysis_one)

##### Recode data for easier analysis.
##### Combine religion and party affiliation values.

trimmed_gss_analysis_one <- trimmed_gss_analysis_one %>% mutate(
    relig = case_match(
      relig,
      "None" ~ "Not Religious",
      .default = "Religious"),
    partyid = case_match(
      partyid,
      "Strong Democrat"~"Democrat",
      "Not Str Democrat"~"Democrat",
      "Ind,Near Dem"~"Democrat",
      "Independent"~"Independent",
      "Ind,Near Rep"~"Republican",
      "Not Str Republican"~"Republican",
      "Strong Republican"~"Republican",
      "Other Party"~"Other"
    ),
    )

##### Rename column headers.
colnames(trimmed_gss_analysis_one) <- c("Religiosity","Political Party")

##### Create contingency table
(partyid_relig_n <- (table(trimmed_gss_analysis_one$Religiosity,
                                   trimmed_gss_analysis_one$'Political Party')))

##### Export contingency table and trimmed data for external Chi-Sq Test of Independence in Excel.
write.csv(partyid_relig_n, "party_religion.csv", row.names = TRUE)
```

#### Plots
##### Mosaic Plot with Residuals
The following plot shows the the association between Political Party affiliation
and religiosity. The data indicate that Democrats, Independents and Other see a
higher proportion of Non-Religious members, in contrast to Republicans which see
the opposite. Based on this, we can determine that political party and religiosity
are dependent on one another.

The standardized residuals indicate the difference between the observed and
expected values.
```{r}
mosaicplot(partyid_relig_n, main='Religiosity by Political Party', 
           col=TRUE, 
           xlab = "Religiosity", 
           ylab="Political Party Affiliation", 
           shade = TRUE,
           off = 4,
           cex.axis = 0.45,
           margin = 1)
```

#### Summary Statistics
##### Contingency Tables
The summary statistics indicate a significant difference between the proportion
of religious and non-religious individuals affiliated with a particular
political party and those that are not

```{r}
(partyid_relig_counts <- (addmargins(table(trimmed_gss_analysis_one$Religiosity,
                                   trimmed_gss_analysis_one$'Political Party'))))

(partyid_relig_prop <- (prop.table(addmargins(table(trimmed_gss_analysis_one$Religiosity,
                                   trimmed_gss_analysis_one$'Political Party')))))
```

* * *

### Part 4: Inference

#### Hypotheses
Does there appear to be a relationship between religiosity and political party 
affiliation?

H0 (nothing is going on) Religion and Party affiliation are independent.
Religiosity does not vary by Political Party.

HA (something is going on) Religion and party affiliation are dependent.
Religiosity does vary by Political Party.

#### Evaluating the Hypothesis
1. Quantify how different the observed counts are from the expected counts.
2. Significant deviations from what is expected based on sampling variation 
(chance) alone provide strong evidence for the alternative hypothesis.

#### Conditions for the Chi-Square Test of Independence

##### Independence:
1. Sampled observations are independent.
  - Random sampling was employed for the survey.

2. n < 10% of the population if sampling without replacement.
  - Observations are less than 10% of the population.

3. Each case only contributes to one cell in the table.
  - It is safe to assume that each individual surveyed reported once, although it's 
not impossible someone took part more than once, and therefore is part of multiple
cells. For the sake of this analysis, we can assume that each case has only 
contributed to one cell.

##### Sample Size:
1. Each cell must have at least 5 expected cases.
  - True.

#### Test the hypothesis 
What is the overall religiosity in the sample?
```{r}
50481/55703
```
Test the hypothesis that religiosity and political party affiliation are 
associated at the 5% significance level.

  - x^2 =	970.29
  - df = 3

```{r}
(chiSq <- pchisq(970.29, 3, lower.tail = FALSE))

if(chiSq < 0.05) {
  print("Reject the null hypothesis")
  } else {
    print("Fail to reject the null hypothesis")
  }
```

#### Conclusion

The p_value of 5.012349e-210 is less than the 0.05 significance level. Therefore,
we reject the null hypothesis in favour of the alternative; there is sufficient
evidence indicating there is an association between Political Party affiliation and
Religiosity.

* * *

## Analysis #2 - Level of respondents education and household income (ANOVA and pairwise tests (theoretical only))

### Part 3: Exploratory data analysis

Does there appear to be a relationship between the level of respondents education
and their household income?

The levels of education contained in this analysis are: Limited HighSchool, 
HighSchool, Junior College, Bachelor, Graduate.

#### Initial Analysis and Cleaning
```{r}
##### Keep specific columns for analysis.
trimmed_gss_analysis_two <- gss %>%
  select(c(coninc, degree))

##### Check for N/A's.
sapply(trimmed_gss_analysis_two, function(x) sum(is.na(x)))

##### Replace missing values with the mean of the column.
#trimmed_gss_analysis_two$coninc[is.na(trimmed_gss_analysis_two$coninc)] <- #mean(trimmed_gss_analysis_two$coninc, na.rm=TRUE)

#### Omit rows with NA's (acceptable due to the large amount of data collected.)
trimmed_gss_analysis_two <- na.omit(trimmed_gss_analysis_two)

##### Rename column headers.
colnames(trimmed_gss_analysis_two) <- c("Family_Income","Degree")

# ##### Recode data for easier analysis.
# trimmed_gss_analysis_two <- trimmed_gss_analysis_two %>% mutate(
#     Degree = case_match(
#       Degree,
#       'Lt High School'~'1',
#       'High School'~'2',
#       'Junior College'~'3',
#       'Bachelor'~'4',
#       'Graduate'~'5',
#     )
# )
               
#### Export data for further analysis in Excel and additional formatting.
write.csv(trimmed_gss_analysis_two, "GSS_FamIncome_Degree_ANOVA.csv", row.names = TRUE)
```

#### Plots
##### Jitter Plot

The graph indicates a clear difference between various levels of respondent education
and associated family earnings. The largest appears to be between Limited High School
and High School; however, there are apparent differences among many of the
degrees of education.

```{r}
ggplot(trimmed_gss_analysis_two) +
  aes(x = Degree, y = Family_Income, color = Degree) +
  geom_jitter() +
  theme(legend.position = "none")
```

* * *

### Part 4: Inference

#### Hypotheses
Does there appear to be a relationship between the level of respondents education
and their household income?

  - H0: The average family income is the same across all degrees of education. (u1 = u2 = u3 = u4 = u5)
  - HA: The average family income differs between at least one pair of degrees of education.

#### Evaluating the Hypothesis

##### Conditions for the ANOVA Test

```{r}
res_aov <- aov(Family_Income ~ Degree,
  data = trimmed_gss_analysis_two
)
```

###### Independence
Within: sampled observations must be independent of each other
1. Random sample / assignment
2. Each nj less than 10% of respective population
3. Always important, but sometimes difficult to check
  - Respondents were sampled randomly as part of the survey, therefore, it is safe to 
assume that the observations are independent of one another.
  - The observations are less than 10% of the respective populations.

Between: groups must be independent of each other
1. Carefully consider whether the groups may be dependent
  - Random sampling was employed, so it's safe to assume that each observation is
independent, and therefore each group is also independent of one another as well
as each observation is contained within one cell.

###### Approximately Normal
1. Distribution of response variable within each group should be 
approximately normal
2. Especially important when sample sizes are small
  - Although there is skew in the data, the sample size is very large, therefore
this shouldn't be an issue.

```{r}
hist(res_aov$residuals)
```

###### Constant Variance

1. Variability should be consistent across groups: homoscedastic groups
2. Especially important when sample sizes differ between groups
  - Variance does not appear to be constant among all groups, therefore an 
alternative ANOVA test is required.

```{r}
ggplot(trimmed_gss_analysis_two) +
  aes(x = Degree, y = Family_Income) +
  geom_boxplot()

qqPlot(res_aov$residuals,
  id = FALSE # id = FALSE to remove point identification
)
```

#### Test the hypothesis - ANOVA

How much variability is attributed to the explanatory variable? (Education)

```{r}
1.08127E+13/6.53791E+13

print("Approximately 16% of variability is due to the explanatory variable.")
```
Test the hypothesis that Family Income and respondent Education are 
associated at the 5% significance level.

```{r}
oneway.test(Family_Income ~ Degree,
  data = trimmed_gss_analysis_two,
  var.equal = FALSE # assuming unequal variances
)
```

```{r}
pf(2340.8,4,10598,lower.tail=FALSE)
```
#### Conclusion
  - If p-value is small (less than α), reject H0. 
  - The data provide convincing evidence that at least 
one pair of population means are different from each 
other (but we can’t tell which one).

  - If p-value is large, fail to reject H0. 
  - The data do not provide convincing evidence that at 
least one pair of population means are different from 
each other, the observed differences in sample means 
are attributable to sampling variability (or chance).

The p_value of 0 is less than the 0.05 significance level. Therefore,
we reject the null hypothesis in favour of the alternative; there is sufficient
evidence indicating that at least one pair of population means are different 
from one another, however, we do not know which.

#### Test the hypothesis - Pairwise Testing

Modified Significance level given 5 levels: 0.02
SE for multiple pairwise comparisons: 382.773
Degrees of Freedom for multiple pairwise comparisons: 10267

1. Is there a difference between the average family income of Lt High School and 
High School respondents?

```{r}
(lHS_HS <- 2 * pt(43.332, df = 10267, lower.tail = FALSE))

if(lHS_HS < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

2. Is there a difference between the average family income of Lt High School and 
Junior College respondents?

```{r}
(lHS_JC <- 2 * pt(35.078, df = 2803, lower.tail = FALSE))

if(lHS_JC < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

3. Is there a difference between the average family income of Lt High School and 
Bachelor respondents?

```{r}
(lHS_B <- 2 * pt(55.011, df = 7373, lower.tail = FALSE))

if(lHS_B < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

4. Is there a difference between the average family income of Lt High School and 
Graduate respondents?

```{r}
(lHS_G <- 2 * pt(-39.771, df = 3565, lower.tail = FALSE))

if(lHS_G < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

5. Is there a difference between the average family income of High School and 
Junior College respondents?

```{r}
(HS_JC <- 2 * pt(12.255, df = 2803, lower.tail = FALSE))

if(HS_JC < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

6. Is there a difference between the average family income of High School
and Bachelor respondents?

```{r}
(HS_B <- 2 * pt(50.723, df = 7373, lower.tail = FALSE))

if(HS_B < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

7. Is there a difference between the average family income of High School
and Graduate respondents?

```{r}
(HS_G <- 2 * pt(61.497, df = 3565, lower.tail = FALSE))

if(HS_G < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

8. Is there a difference between the average family income of Junior College
and Bachelor respondents?

```{r}
(JC_B <- 2 * pt(19.144, df = 2803, lower.tail = FALSE))

if(JC_B < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

9. Is there a difference between the average family income of Junior College
and Graduate respondents?

```{r}
(JC_G <- 2 * pt(33.828, df = 2803, lower.tail = FALSE))

if(JC_G < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

10. Is there a difference between the average family income of Bachelor
and Graduate respondents?

```{r}
(B_G <- 2 * pt(21.036, df = 3565, lower.tail = FALSE))

if(B_G < 0.02) {
  print("Reject the null hypothesis")
} else {
  print("Fail to reject the null hypothesis")
}
```

#### Conclusion
All pairs but Lt High School vs Graduate showed a significant difference.

## Analysis #3 - Average # of Children

How many children do respondents in this sample have on average?
Estimate the true, average number of children based on this sample with a 95%
confidence interval.

* * *

### Part 3: Exploratory data analysis

```{r}
##### Keep specific columns for analysis.
trimmed_gss_analysis_three <- gss %>%
  select(c(childs))

##### Check for N/A's.
sapply(trimmed_gss_analysis_three, function(x) sum(is.na(x)))

##### Replace missing values with the mean of the column.
#trimmed_gss_analysis_two$coninc[is.na(trimmed_gss_analysis_two$coninc)] <- #mean(trimmed_gss_analysis_two$coninc, na.rm=TRUE)

#### Omit rows with NA's (acceptable due to the large amount of data collected.)
trimmed_gss_analysis_three <- na.omit(trimmed_gss_analysis_three)

##### Rename column headers.
colnames(trimmed_gss_analysis_three) <- c("Children")
               
#### Export data for further analysis in Excel, if necessary.
write.csv(trimmed_gss_analysis_three, "c_data.csv", row.names = TRUE)
```

#### Plots
##### Boxplot and Barplot

```{r}
ggplot(trimmed_gss_analysis_three) +
  geom_boxplot(fill = "lightblue", color = "darkblue", size = 1, outlier.color = "red", outlier.shape = 8) +
  aes(y = Children)

ggplot(trimmed_gss_analysis_three) +
  geom_bar(fill = "lightblue", color = "darkblue", linewidth = 1) +
  aes(y = Children)
```
#### Summary Statistics

```{r}
summary(trimmed_gss_analysis_three)
```
* * *

### Part 4: Inference

#### Conditions

##### Independence
1. Independence: Sampled observations must be independent. 
2. random sample/assignment
3. if sampling without replacement, n < 10% of population
  - Random sample & 56,880 < 10% of the population at large.
  - Independence cannot be guaranteed, as there is a chance multiple respondents
from the same household were surveyed, or individuals who are no longer in the
same household provide data on the same children.

##### Approximately Normal
1. Sample size/skew: n ≥ 30, larger if the population distribution is very skewed.
  - Sample size/skew: 56,880 ≥ 30.
  - There is slight skew in the data, however, the sample size is large.

#### Calculate the Confidence Interval 

```{r}
#### Calculate Mean, Median and Standard Deviation.
(trimmed_gss_analysis_three %>%
  summarize(Mean=mean(Children), Median=median(Children),standard_deviation=sd(Children)))

se = sd(trimmed_gss_analysis_three$Children/(sqrt(56880)))

CI_low = mean(trimmed_gss_analysis_three$Children - 1.96*se)
CI_high = mean(trimmed_gss_analysis_three$Children + 1.96*se)

(CI <- c(CI_low,CI_high))
```

#### Conclusion

We are 95% confident that adults, on average, have approximately 1.94 - 1.97
children.

#### Test the hypothesis

A 95% confidence interval for the average number of children adults have was 
(1.94, 1.97). Based on this confidence interval, do these data 
support the hypothesis that adults on average have more than 1.95 children?
(In this case, the sample mean.)

H0: μ = 1.95: Adults have 1.95 children on average.
HA: μ > 1.95: Adults have more than 1.95 children on average.

p-value: P(observed or more extreme outcome | H0 true)
P(X > 1.95 | H0: μ = 1.95)
X ~ N(μ = 1.95, SE = 0.0075)

  - N = 56880
  - x_bar = 1.95
  - SD = 1.79
  - SE = 0.0075

```{r}
pnorm <- pnorm(((mean(trimmed_gss_analysis_three$Children - 1.95)/sd(trimmed_gss_analysis_three$Children/(sqrt(56880))))),lower.tail=FALSE)

if(pnorm < 0.05) {
  print("Reject the null hypothesis")
  } else {
    print("Fail to reject the null hypothesis")
  }
```
#### Conclusion

Since the p_value is high (greater than 0.05), we fail to reject the null hypothesis. 
There is insufficient data indicating that adults have more than 1.95 children on average.

## Analysis #4 - Average age of adults in the US

### Part 3: Exploratory data analysis

What is the average age of the US adult population?
Estimate the true, average age of adults in the US, based on this sample with a 
95% confidence interval.

```{r}
##### Keep specific columns for analysis.
trimmed_gss_analysis_four <- gss %>%
  select(c(age))

##### Check for N/A's.
sapply(trimmed_gss_analysis_four, function(x) sum(is.na(x)))

##### Replace missing values with the mean of the column.
#trimmed_gss_analysis_two$coninc[is.na(trimmed_gss_analysis_two$coninc)] <- #mean(trimmed_gss_analysis_two$coninc, na.rm=TRUE)

#### Omit rows with NA's (acceptable due to the large amount of data collected.)
trimmed_gss_analysis_four <- na.omit(trimmed_gss_analysis_four)

##### Rename column headers.
colnames(trimmed_gss_analysis_four) <- c("Age")
               
#### Export data for further analysis in Excel, if necessary.
write.csv(trimmed_gss_analysis_four, "a_data.csv", row.names = TRUE)
```

#### Plots
##### Boxplot and Barplot

```{r}
ggplot(trimmed_gss_analysis_four) +
  geom_boxplot(fill = "lightblue", color = "darkblue", size = 1, outlier.color = "red", outlier.shape = 8) +
  aes(y = Age)

ggplot(trimmed_gss_analysis_four) +
  geom_bar(fill = "lightblue", color = "darkblue", linewidth = 1) +
  aes(y = Age)
```

#### Summary Statistics

```{r}
summary(trimmed_gss_analysis_four)
```

### Part 4: Inference

#### Conditions

##### Independence
1. Independence: Sampled observations must be independent. 
2. random sample/assignment
3. if sampling without replacement, n < 10% of population
  - Random sample & 56,859 < 10% of the adult population at large.
  - One adult is surveyed per home. Being that this analysis is on the average
  age of adults, even if, somehow, more than one adult from the same household
  provided a response, it wouldn't drastically impact the integrity of the 
  analysis.

##### Approximately Normal
1. Sample size/skew: n ≥ 30, larger if the population distribution is very skewed.
  - Sample size/skew: 56,859 ≥ 30.
  - There is slight skew in the data, however, the sample size is large.

#### Calculate the Confidence Interval 

```{r}
#### Calculate Mean, Median and Standard Deviation.
(trimmed_gss_analysis_four %>%
  summarize(Mean=mean(Age), Median=median(Age),standard_deviation=sd(Age)))

se = sd(trimmed_gss_analysis_four$Age/(sqrt(56859)))

CI_low = mean(trimmed_gss_analysis_four$Age - 1.96*se)
CI_high = mean(trimmed_gss_analysis_four$Age + 1.96*se)

(CI <- c(CI_low, CI_high))
```

We are 95% confident that the average adult age in the US is between 45.55 and
45.84 years.

#### Test the hypothesis

A 95% confidence interval for the average age of adults have was 
(45.55434 45.84157). Based on this confidence interval, do these data 
support the hypothesis that the average age of Adults in the US is greater than
43 (the median)?

  - H0: μ = 43: Average age of Adults in the US is 43.
  - HA: μ > 43: Average age of Adults in the US is greater than 43.

p-value: P(observed or more extreme outcome | H0 true)
P(X > 45.7 | H0: μ = 43)
X ~ N(μ = 43, SE = 0.0075)

  - N = 56859
  - x_bar = 45.7
  - SD = 17.47211
  - SE = 0.07327331
  
```{r}
pnorm <- pnorm(((mean(trimmed_gss_analysis_four$Age - 43)/sd(trimmed_gss_analysis_four$Age/(sqrt(56859))))),lower.tail=FALSE)

if(pnorm < 0.05) {
  print("Reject the null hypothesis")
  } else {
    print("Fail to reject the null hypothesis")
  }
```
#### Conclusion

Since the p_value is high (greater than 0.05), we reject the null hypothesis.
There is sufficient data indicating that the average age of Adults in the Us
is greater than 43.